Non-Monotone GLS Simulation: Simple

### 2024-02-02 src/explore/20240202-gls_sim.qmd
Changed project structure.

---
title: "Non-Monotone GLS Simulation: No $A_{00}$"
author:
  - name: Caleb Leedy
date: today
date-format: "D MMMM YYYY"
format:
  pdf:
    include-in-header: 
      - "latex_header.tex"
    keep-tex: true
---

@@

  # A_11
  x <- rnorm(n_obs_sec)
  e_mat <- MASS::mvrnorm(n_obs_sec, c(0, 0), matrix(c(1, rho, rho, 1), nrow = 2))
  e1 <- e_mat[, 1]
  e2 <- e_mat[, 2]
  y1 <- x + e1
  y2 <- mu + x + e1

  df_11 <- 
    tibble(X = x, Y1 = y1, Y2 = y2,
           delta_00 = 0, delta_10 = 0, delta_01 = 0, delta_11 = 1,
           pi_00 = 1/3, pi_10 = 1/3, pi_01 = 1/3, pi_11 = 1/3)
  # A_10
  x <- rnorm(n_obs_sec)
  e_mat <- MASS::mvrnorm(n_obs_sec, c(0, 0), matrix(c(1, rho, rho, 1), nrow = 2))
  e1 <- e_mat[, 1]
  e2 <- e_mat[, 2]
  y1 <- x + e1
  y2 <- mu + x + e1

  df_10 <- 
    tibble(X = x, Y1 = y1, Y2 = y2,
           delta_00 = 0, delta_10 = 1, delta_01 = 0, delta_11 = 0,
           pi_00 = 1/3, pi_10 = 1/3, pi_01 = 1/3, pi_11 = 1/3)
  # A_01
  x <- rnorm(n_obs_sec)
  e_mat <- MASS::mvrnorm(n_obs_sec, c(0, 0), matrix(c(1, rho, rho, 1), nrow = 2))
  e1 <- e_mat[, 1]
  e2 <- e_mat[, 2]
  y1 <- x + e1
  y2 <- mu + x + e1

  df_01 <- 
    tibble(X = x, Y1 = y1, Y2 = y2,
           delta_00 = 0, delta_10 = 0, delta_01 = 1, delta_11 = 0,
           pi_00 = 1/3, pi_10 = 1/3, pi_01 = 1/3, pi_11 = 1/3)
  # A_00
  x <- rnorm(n_obs_sec)
  e_mat <- MASS::mvrnorm(n_obs_sec, c(0, 0), matrix(c(1, rho, rho, 1), nrow = 2))
  e1 <- e_mat[, 1]
  e2 <- e_mat[, 2]
  y1 <- x + e1
  y2 <- mu + x + e1

  df_00 <- 
    tibble(X = x, Y1 = y1, Y2 = y2,
           delta_00 = 1, delta_10 = 0, delta_01 = 0, delta_11 = 0,
           pi_00 = 1/3, pi_10 = 1/3, pi_01 = 1/3, pi_11 = 1/3)

  df <- bind_rows(df_11, df_10, df_01)

@@

  if (class == "basic") {
    z_mat <- matrix(c(1, 0, 0,
                      0, 1, 0,
                      0, 0, 1,
                      1, 0, 0,
                      0, 1, 0,
                      1, 0, 0,
                      0, 0, 1), ncol = 3, byrow = TRUE)

    g_hat <- 
      c(mutate(df, tmp = delta_11 / pi_11 * X) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_11 / pi_11 * Y1) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_11 / pi_11 * Y2) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_10 / pi_10 * X) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_10 / pi_10 * Y1) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_01 / pi_01 * X) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_01 / pi_01 * Y2) |> pull(tmp) |> mean())

    v_11 <- 1 / df$pi_11[1] *
      matrix(c(1, 1, 1,
               1, 2, 1 + rho,
               1, 1 + rho, mu^2 + 2), nrow = 3, byrow = TRUE) -
      matrix(c(0, 0, 0,
               0, 0, 0,
               0, 0, mu^2), nrow = 3, byrow = TRUE)
    v_10 <- 1 / df$pi_10[1] * matrix(c(1, 1, 1, 2), nrow = 2, byrow = TRUE)
    v_01 <- 
      1 / df$pi_01[1] * matrix(c(1, 1, 1, mu^2 + 2), nrow = 2) - 
      matrix(c(0, 0, 0, mu^2), nrow = 2)

    v_mat <- matrix(rep(0, 49), ncol = 7)
    v_mat[1:3, 1:3] <- v_11
    v_mat[4:5, 4:5] <- v_10
    v_mat[6:7, 6:7] <- v_01

@@

  } else {
    stop("We have only 1 class: 'basic'.")
  }
}

```{r}
#| label: MCMC
#| echo: false
#| tbl-cap: |
#|   Results from simulations study with independent equally sized segments
#|   $A_{11}$, $A_{10}$, and $A_{01}$ all of size $n = 250$. In this
#|   simulation we have the true mean of $Y_2$ equal to $\mu = 5$ and the covariance
#|   between $e_1$ and $e_2$ is $\rho = 0.5$. The goal is to estimate 
#|   $E[Y_2] = \mu$. For the GLS estimation, we use the
#|   true known covariance matrix (using the true values of $\mu$ and $\rho$),
#|   and we use g-functions $g_1 = X$, $g_2 = Y_1$ and $g_3 = Y_2$.

@@

# Analyze Results
mc_theta |>
  summarize(
    across(everything(),
    list(Bias = ~mean(.) - true_mu,
         SD = sd,
         Tstat = ~(mean(.) - true_mu) / sqrt(var(.) / B)),
    .names = "{.fn}_{.col}")) |> 
  tidyr::pivot_longer(cols = everything(),
                      names_to = c(".value", "Algorithm"),
                      names_pattern = "(.*)_(.*)") |>
  mutate(Pval = pt(-abs(Tstat), df = B)) |>
  knitr::kable(digits = 3)

@@

A quick comparison between the standard deviation of the GLS estimator and the
standard deviation of the GLS estimator in `gls_sim.qmd` shows that this
standard deviation is slightly larger, which means that we do lose efficiency by
not including $A_{00}$.

>>>>

---
title: "Non-Monotone GLS Simulation: Simple"
author:
  - name: Caleb Leedy
date: today
date-format: "D MMMM YYYY"
format:
  pdf:
    include-in-header: 
      - "latex_header.tex"
    keep-tex: true
---

@@

  # A_11
  x <- rnorm(n_obs_sec)
  e_mat <- MASS::mvrnorm(n_obs_sec, c(0, 0), matrix(c(1, rho, rho, 1), nrow = 2))
  e1 <- e_mat[, 1]
  e2 <- e_mat[, 2]
  y1 <- x + e1
  y2 <- mu + x + e2

  df_11 <- 
    tibble(X = x, Y1 = y1, Y2 = y2,
           delta_00 = 0, delta_10 = 0, delta_01 = 0, delta_11 = 1,
           pi_00 = 0.25, pi_10 = 0.25, pi_01 = 0.25, pi_11 = 0.25)
  # A_10
  x <- rnorm(n_obs_sec)
  e_mat <- MASS::mvrnorm(n_obs_sec, c(0, 0), matrix(c(1, rho, rho, 1), nrow = 2))
  e1 <- e_mat[, 1]
  e2 <- e_mat[, 2]
  y1 <- x + e1
  y2 <- mu + x + e2

  df_10 <- 
    tibble(X = x, Y1 = y1, Y2 = y2,
           delta_00 = 0, delta_10 = 1, delta_01 = 0, delta_11 = 0,
           pi_00 = 0.25, pi_10 = 0.25, pi_01 = 0.25, pi_11 = 0.25)
  # A_01
  x <- rnorm(n_obs_sec)
  e_mat <- MASS::mvrnorm(n_obs_sec, c(0, 0), matrix(c(1, rho, rho, 1), nrow = 2))
  e1 <- e_mat[, 1]
  e2 <- e_mat[, 2]
  y1 <- x + e1
  y2 <- mu + x + e2

  df_01 <- 
    tibble(X = x, Y1 = y1, Y2 = y2,
           delta_00 = 0, delta_10 = 0, delta_01 = 1, delta_11 = 0,
           pi_00 = 0.25, pi_10 = 0.25, pi_01 = 0.25, pi_11 = 0.25)
  # A_00
  x <- rnorm(n_obs_sec)
  e_mat <- MASS::mvrnorm(n_obs_sec, c(0, 0), matrix(c(1, rho, rho, 1), nrow = 2))
  e1 <- e_mat[, 1]
  e2 <- e_mat[, 2]
  y1 <- x + e1
  y2 <- mu + x + e2

  df_00 <- 
    tibble(X = x, Y1 = y1, Y2 = y2,
           delta_00 = 1, delta_10 = 0, delta_01 = 0, delta_11 = 0,
           pi_00 = 0.25, pi_10 = 0.25, pi_01 = 0.25, pi_11 = 0.25)

  df <- bind_rows(df_11, df_10, df_01, df_00)

@@

  if (class == "basic") {
    z_mat <- matrix(c(1, 0, 0,
                      0, 1, 0,
                      0, 0, 1,
                      1, 0, 0,
                      0, 1, 0,
                      1, 0, 0,
                      0, 0, 1,
                      1, 0, 0), ncol = 3, byrow = TRUE)

    g_hat <- 
      c(mutate(df, tmp = delta_11 / pi_11 * X) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_11 / pi_11 * Y1) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_11 / pi_11 * Y2) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_10 / pi_10 * X) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_10 / pi_10 * Y1) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_01 / pi_01 * X) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_01 / pi_01 * Y2) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_00 / pi_00 * X) |> pull(tmp) |> mean())

    v_11 <- 1 / df$pi_11[1] *
      matrix(c(1, 1, 1,
               1, 2, 1 + rho,
               1, 1 + rho, 2), nrow = 3)
    v_10 <- 1 / df$pi_10[1] * matrix(c(1, 1, 1, 2), nrow = 2)
    v_01 <- 1 / df$pi_01[1] * matrix(c(1, 1, 1, 2), nrow = 2)
    v_00 <- 1 / df$pi_00[1]

    v_mat <- matrix(rep(0, 64), ncol = 8)
    v_mat[1:3, 1:3] <- v_11
    v_mat[4:5, 4:5] <- v_10
    v_mat[6:7, 6:7] <- v_01
    v_mat[8, 8] <- v_00

@@

  } else {
    stop("We have only 1 class: 'basic'.")
  }
}
```

```{r}
#| label: MCMC
#| echo: false
#| tbl-cap: |
#|   Results from simulations study with independent equally sized segments
#|   $A_{11}$, $A_{10}$, $A_{01}$, and $A_{00}$ all of size $n = 250$. In this
#|   simulation we have the true mean of $Y_2$ equal to $\mu = 5$ and the covariance
#|   between $e_1$ and $e_2$ is $\rho = 0.5$. The goal is to estimate 
#|   $E[Y_2] = \mu$. For the GLS estimation, we use the
#|   true known covariance matrix (using the true values of $\mu$ and $\rho$),
#|   and we use g-functions $g_1 = X$, $g_2 = Y_1$ and $g_3 = Y_2$.

@@

# Analyze Results
mc_theta |>
  summarize(
    across(everything(),
    list(Bias = ~mean(.) - true_mu,
         SD = ~sd(.),
         Tstat = ~(mean(.) - true_mu) / sqrt(var(.) / B)),
    .names = "{.fn}_{.col}")) |> 
  tidyr::pivot_longer(cols = everything(),
                      names_to = c(".value", "Algorithm"),
                      names_pattern = "(.*)_(.*)") |>
  mutate(Pval = pt(-abs(Tstat), df = B)) |>
  knitr::kable(digits = 3)

@@

The theoretical variance of $\hat g_{GLS}$ is

```{r}
#| echo: false

z_mat <- matrix(c(1, 0, 0,
                  0, 1, 0,
                  0, 0, 1,
                  1, 0, 0,
                  0, 1, 0,
                  1, 0, 0,
                  0, 0, 1,
                  1, 0, 0), ncol = 3, byrow = TRUE)

v_11 <- 1 / 0.25 *
  matrix(c(1, 1, 1,
           1, 2, 1 + cov_e1e2,
           1, 1 + cov_e1e2, 2), nrow = 3)
v_10 <- 1 / 0.25 * matrix(c(1, 1, 1, 2), nrow = 2)
v_01 <- 1 / 0.25 * matrix(c(1, 1, 1, 2), nrow = 2)
v_00 <- 1 / 0.25

v_mat <- matrix(rep(0, 64), ncol = 8)
v_mat[1:3, 1:3] <- v_11
v_mat[4:5, 4:5] <- v_10
v_mat[6:7, 6:7] <- v_01
v_mat[8, 8] <- v_00

v_mat <- v_mat / (4 * obs_seg)
ones_v <- t(c(0, 0, 1))

theo_var <- 
  ones_v %*% MASS::ginv(t(z_mat) %*% MASS::ginv(v_mat) %*% z_mat) %*% t(ones_v)

$$ V(\hat g_{GLS}) = [0, 0, 1] (Z'V^{-1}Z)^{-1} [0, 0, 1]'.$$

This is equal to `{r} theo_var` which means that the standard deviation 
is `{r} sqrt(theo_var)`.
